Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-03-30 Write-ups/Manuscript.Rmd
data application use daily weigh-in data

---
title: "Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis"
author:
  - Ying Jin
  - Andrew Leroux
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
fontfamily: mathpazo
fontsize: 11pt
geometry: margin = 1in
bibliography: refs.bib
nocite: '@*'
link-citations: yes
linkcolor: blue
csl: ASA.cls
header-includes:
  - \usepackage{setspace}\doublespacing

---
<!-- Abstract:  -->
<!-- ================================================================================================================================= -->
<!--   Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, referred to as dynamic prediction. Traditional methods are often limited in flexibility and computationally intensive, especially with non-Gaussian data. To address these issues, we propose a novel method for dynamic prediction based on Generalized Functional Principal Component Analysis (FPCA). Assume the observed outcome follows an exponential family distribution parameterized by a latent Gaussian function, the proposed method consists of the following steps: 1) Bin the data across functional domain into small, equal-length intervals; 2) Fit local generalized mixed models at every bin to estimate individual latent functions; 3) Fit FPCA model to smooth latent functions and 4) Obtain estimates of subject-specific PC scores using partial observations and recover the unobserved part on the binned grid. Our simulation study showed the proposed method achieved significantly better out-of-sample predictive performance compared to existing methods with much shorter computation time, thus has the potential to be widely applicable to large datasets. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(here)
library(ROCR)
library(kableExtra)
library(knitr)
options(knitr.kable.NA = "")

@@

load(here("Data/sim_data.RData"))
load(here("Data/SimOutput_fGFPCA.RData"))
load(here("Data/SimOutput_GLMMadaptive.RData"))
```

@@

<!-- -	Overview of dynamic prediction methods -->

@@

To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain. Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function, and non-parametric smoothing [@Scheipl2014] can be incorporated to speed up the computation, such as spline basis functions or eigenfunctions from functional principal component analysis (fPCA). The introduction of basis functions also makes out-of-sample prediction more straightforward. Instead of estimating subject-specific random effects of new observations, we can simply estimate coefficients/loadings on the basis function used for smoothing. Existing research on dynamic prediction with functional data analysis methods has been focusing on continuous/Gaussian outcomes, modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017]. @kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains. While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors. However, little extension was made on prediction of non-Gaussian functions, such as binary and count outcomes.

@@

In this paper, we aim to develop a fast, scalable method for dynamic prediction of discrete function tracks based on functional mixed effect model with fPCA smoothing. Section 2 presents the procedure of the proposed method. In Section 3, we illustrate the performance and efficiency of our proposed method in a simulation study. In Section 4, we apply this method to a real-world dataset. Section 5 presents a discussion of davantages and limitation of the proposed method.

@@

where g is a appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function follows a zero-mean Gaussian process.

While $b_i(t)$ is not observed, it can be approximated under the FPCA framework $b_i(t)=\sum_{k=1}^K\xi_{ik}\phi_{ik}(t)+\epsilon_i(t)$. Here $\phi_{ik}(t)$ is a set of orthogonal eigenfunctions that explains the most variation in $b_i(t)$, and $\xi_{ik}$ are subject-specific PC scores (or loadings) on each eigenfunction. Additionally, $\xi_{ik}$ are mutually independent obver both subject ($i$) and eigenfunctions ($k$). That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$ where $\lambda_k$ is the kth eigenvalues: $\int \phi_k^2(t)dt=\lambda_k$. $\epsilon_i(t)$ here is a residual function that accounts for the variation not explained by the first K eigenfunctions from FPCA. We assume it follows a zero-mean Gaussian process. At a specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.

Based on the problem set up above, we propose the following algorithm for PFCA on the unobserved latent process $\eta_i(t)$:

1. Bin the observed outcomes in to small, non-overlapping, equal length intervals. We hereafter index the bins by their midpoints $s$. 
2. Fit a local Generalized Mixed Model at every bin. Specifically, at bin $s$, we fit $g[E(Y_i(t))] = \beta_0(s)+b_i(s)$ for all $t$ in bin $s$. From this series of models we can get estiamtes of population mean $\hat{\beta}_0(s)$ and subject-level random effect $\hat{b}_i(s)$, thus estimate of the individual latent functions at every bin: $\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)$. 
3. Fit FPCA on the estimated latent functions $\hat{\eta}_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, population mean $\hat{\beta}_0(s)$ and residual variance $\hat{\sigma}^2$. 
4. With components extracted above, calculate the maximum likelihood estimate (MLE) of the subject-specific PC scores $\hat{\xi}_{ik}$ of new samples based on their partially observed data. Then the value of latent functions at unobserved points can be estimated as $\hat{\eta_i}(s)=\hat{\beta}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)$

Following the algorithm above, predictions of individual latent functions are made on the binned grid based on partially observed non-Gaussian functions tracks of new subjects. Since the bins are set up to be small in length, the binned grid would still be dense enough. However, in situations where we need predictions on the original, un-binned grid instead, linear interpolation turns out to be a fast, convenient way with good performance for prediction at points between bins.

Precision of prediction, usually quantified by the variance of prediction error $Var(\hat{\eta_i}(s)-\eta_i(s))$, is also straightforward under this framework. In step 4 we calculated the MLE of $\hat{\xi}_{ik}$. Based on likelihood theory, its variance can be estimated with observed information ${I}(\hat{\xi}_{ik})$, which is essentially the second derivative of likelihood at  $\hat{\xi}_{ik}$. Therefore, the variation of prediction interval is:

$$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$
Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$.

<!-- -	Need better notation system -->
<!-- - Change to general exponential family -->
<!-- - Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate -->

@@

In this section, we illustrate the predictive performance and computational efficiency of the proposed method through a simulation study. We simulated 50 datasets, each with 500 subjects. For every subject, we generate 1000 binary outcomes $Y_i(t) \in (0, 1)$ across functional domain $t \in [0, 1]$, where the distribution of outcome is characterized by a continuous latent function. The data generation mechanism can be expressed as follows:

$$\begin{aligned}
Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)
\end{aligned}$$

In this simulation, we set $f_0(t) = 0$. $\xi_{ik}$ are mutually independent normal random variables $\xi_{ik}\sim N(0, \gamma_k)$. Here we set the values of $\gamma_k$ to be $0.5^{k-1}$, $k \in (1,..., 4)$. In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $[0, 1]$ and are the same for all subjects.

We use two metrics to evaluate the out-of-sample predictive performance: integrated squared error (ISE) and Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction accucay of latent continuous function. It is evaluated on the binned grid at midpoints of each unobserved bin. If the entire functional domian has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$. The second metric, AUC, focuses on evaluation of prediction of the binary outcome. Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid as well and estimated values of latent functions between bins with linear interpolation.

@@

The average ISE and AUC across all simulation is presented in Table 1, and Figure 1 presents the predicted function curves of four randomly-drawn subjects from the first simulation. The prediction is conditioning on different length of observed track (specifically, with observations up to t = 200, 400, 600, 800 respectively), and evaluation is made on equal-length time windows on the unobserved tracks following the maximum observation time. As Table 1 reveals, the fGFPCA outperformed GLMMadaptive under every scenario, which was expected since GLMMadaptive can only accomodate simple model structures. In addition, fGFPCA also takes less computation time. For one simulated dataset, fGFPCA spent `r round(mean(runtime),2)` minutes on model fitting and out-of-sample prediction, while GLMMadaptive took `r round(mean(runtime_ref),2)` minutes.

@@

```{r}

# other quantities
mid <- unique(pred_lst[[1]]$bin)
N <- 500
M <- length(pred_lst)
J <- 1000

@@

```{r Figure, fig.cap="Predicted track of four randomly selected subjects. The grey dash line indicates the true latent continuous function. The dashed lines indicated predicted latent function tracks, and color indicates different observations time."}
# figure of fGFPCA prediction
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

@@

@@

exp_df_sim1 %>%
  left_join(df_true_sim1 %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_i, col = "True"))+
  geom_line(aes(x=bin, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Bin", colour= "Observed time")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))
```

@@

Discussion
=================================================================================================================================
-	Grid 
-	Score bias: cannot demonstrate without repeat simulation

>>>>

---
title: "Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis"
author:
  - Ying Jin
  - Andrew Leroux
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
fontfamily: mathpazo
fontsize: 11pt
geometry: margin = 1in
bibliography: refs.bib
nocite: '@*'
link-citations: yes
linkcolor: blue
csl: ASA.cls
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage[utf8]{inputenc}
  - \usepackage[T1]{fontenc}
---
<!-- Abstract:  -->
<!-- ================================================================================================================================= -->
<!--   Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, referred to as dynamic prediction. Traditional methods are often limited in flexibility and computationally intensive, especially with non-Gaussian data. To address these issues, we propose a novel method for dynamic prediction based on Generalized Functional Principal Component Analysis (FPCA). Assume the observed outcome follows an exponential family distribution parameterized by a latent Gaussian function, the proposed method consists of the following steps: 1) Bin the data across functional domain into small, equal-length intervals; 2) Fit local generalized mixed models at every bin to estimate individual latent functions; 3) Fit FPCA model to smooth latent functions and 4) Obtain estimates of subject-specific PC scores using partial observations and recover the unobserved part on the binned grid. Our simulation study showed the proposed method achieved significantly better out-of-sample predictive performance compared to existing methods with much shorter computation time, thus has the potential to be widely applicable to large datasets. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(here)
library(ROCR)
library(kableExtra)
library(knitr)
library(ggpubr)
options(knitr.kable.NA = "")

@@

# simulation
load(here("Data/sim_data.RData"))
load(here("Data/SimOutput_fGFPCA.RData"))
load(here("Data/SimOutput_GLMMadaptive.RData"))

# data application
df_appl <- read.csv(here("Data/Daily_Weigh_ins.csv"))
load(here("Data/ApplOutput_fGFPCA.RData"))
load(here("Data/ApplOutput_GLMMadaptive.RData"))

@@

@@

To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain of measures. Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function, and non-parametric smoothing [@Scheipl2014] can be incorporated to speed up the computation, such as spline basis functions or eigenfunctions from functional principal component analysis (FPCA). The introduction of basis functions also makes out-of-sample prediction more straightforward. Instead of estimating subject-specific random effects of new observations, we can simply estimate coefficients/loadings on the basis function used for smoothing. Existing research on dynamic prediction with functional data analysis methods has been focusing on continuous/Gaussian outcomes, modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017]. @kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains. While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors.

@@

In this paper, we aim to develop a fast, scalable method for dynamic prediction of discrete function tracks based on functional mixed effect model with FPCA smoothing. Section 2 presents the procedure of the proposed method. In Section 3, we illustrate the performance and efficiency of our proposed method in a simulation study. In Section 4, we apply this method to a real-world dataset. Section 5 presents a discussion of advantages and limitation of the proposed method.

@@

where g is an appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function following a zero-mean Gaussian process.

According to the Karhunen-Loève theorem, the unobserved random effect function $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$. We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)$, where $\epsilon_i(t)$ indicates random noise introduced by this approximation, following a zero-mean Gaussian process. At each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.

In practice, this approximation could be implemented through an FPCA procedure on $\eta_i(t)$. In this case $\phi_{k}(t),\ k \in \{1,...,K\}$ are the first K eigenfunctions that explain the most variation in $\eta_i(t)$, and $\xi_{ik}$ are subject-specific PC scores (or loadings) on each eigenfunction. Additionally, $\xi_{ik}$ are mutually independent over both subject ($i$) and eigenfunctions ($k$). That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$ where $\lambda_k$ is the kth eigenvalues.

**Fast Generalized FPCA** Based on the problem set up above, we propose the following algorithm for fast, generalizable implementation of FPCA (fGFPCA) on the unobserved latent process $\eta_i(t)$:

1. Bin the observed outcomes in to small, non-overlapping, equal length intervals. We hereafter index the bins by their midpoints $s$. 
2. Fit a local Generalized Mixed Model at every bin. Specifically, at bin $s$, we fit $g[E(Y_i(t))] = \beta_0(s)+b_i(s)$ for all $t$ in bin $s$. From this series of models we can estimate the individual latent functions at every bin: $\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)$. 
3. Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.

**Dynamic prediction** Now assume we have a new observations with partially observed outcome $Y_{new}(t)$, $t\leq t_m$. The maximum observed point is $t_m$, which is a point along the functional domain and belongs to bin $s_m$ on the binned grid from step 1 above. With components extracted from the FPCA model in step 3, we will be able to make predictions on the future, unobserved bins. Specifically, with the partially observed binary function $Y_{new}(t)$, we will be able to calculate the maximum likelihood estimate (MLE) of the subject-specific PC scores $\hat{\xi}_{ik}$. Then the value of latent functions at unobserved points can be estimated as $\hat{\eta_i}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)$

Following the algorithm above, predictions of individual latent functions are made on the binned grid based on partially observed non-Gaussian functions tracks of new subjects. Since the bins are set up to be small in length, the binned grid would still be dense. However, in situations where predictions on the original, un-binned grid is needed, linear interpolation turns out to be a fast, convenient way with good performance for prediction at points between bins.

<!-- Precision of prediction, usually quantified by the variance of prediction error $Var(\hat{\eta_i}(s)-\eta_i(s))$, is also straightforward under this framework. In step 4 we calculated the MLE of $\hat{\xi}_{ik}$. Based on likelihood theory, its variance can be estimated with observed information ${I}(\hat{\xi}_{ik})$, which is essentially the second derivative of likelihood at  $\hat{\xi}_{ik}$. Therefore, the variation of prediction interval is: -->

<!-- $$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$ -->
<!-- Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$. -->

<!-- - Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate -->

@@

In this section, we illustrate the predictive performance and computational efficiency of the proposed method through a simulation study. We simulated 50 datasets, each with 500 subjects. For every subject, we generate 1000 binary outcomes $Y_i(t) \in \{0, 1\}$ across functional domain $t \in [0, 1000]$, where the distribution of outcome is characterized by a continuous latent function. The data generation mechanism can be expressed as follows:

$$\begin{aligned}
Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
\end{aligned}$$

In this simulation, we set $f_0(t) = 0$. $\xi_{ik}$ are mutually independent normal random variables $\xi_{ik}\sim N(0, \gamma_k)$. Here we set the values of $\gamma_k$ to be $0.5^{k-1}$, $k \in (1,..., 4)$. In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $[0, 1000]$ and are the same for all subjects.

We use two metrics to evaluate the out-of-sample predictive performance: integrated squared error (ISE) and Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction accuracy of latent continuous function. It is evaluated on the binned grid at midpoints of each unobserved bin. If the entire functional domain has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=s_m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$. The second metric, AUC, focuses on evaluation of prediction of the binary outcome. Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid and estimated values of latent functions between bins with linear interpolation.

@@

The average ISE and AUC across all simulation is presented in Table 1, and Figure 1 presents the predicted function curves of four randomly-drawn subjects from the first simulation. The prediction is conditioning on different length of observed track (specifically, with observations up to t = 200, 400, 600, 800 respectively), and evaluation is made on equal-length time windows on the unobserved tracks following the maximum observation time. As Table 1 reveals, our proposed method (fGFPCA) outperformed GLMMadaptive under every scenario, which was expected since GLMMadaptive can only accommodate simple model structures. In addition, fGFPCA also took less computation time. For one simulated dataset, fGFPCA spent `r round(mean(runtime),2)` minutes on model fitting and out-of-sample prediction, while GLMMadaptive took `r round(mean(runtime_ref),2)` minutes.

@@

```{r}
# other quantities
mid <- unique(pred_lst[[1]]$bin)
N <- 500
M <- length(pred_lst)
J <- 1000

@@

```{r Figure, fig.cap="Predicted track of four randomly selected subjects from one simulated dataset. The grey solid line indicates the true latent continuous function. The dashed lines indicated predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
# figure of fGFPCA prediction
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

@@

exp_df_sim1_ref <- pred_lst_ref[[1]]

@@

# fGFPCA
fig1a <- exp_df_sim1 %>%
  left_join(df_true_sim1 %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_i, col = "True"))+
  geom_line(aes(x=bin, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "fGFPCA")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))

# GLMMadaptive
fig1b <- exp_df_sim1_ref %>%
  filter(sind_inx %in% mid) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=sind_inx, y=eta_i, col = "True"))+
  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "GLMMadaptive")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))

ggarrange(fig1a, fig1b, nrow = 1, common.legend = T)

@@

```{r}
# data clean
mid2 <- unique(df_est_latent$bin)
N2 <- max(df_est_latent$id)
J2 <- 400

# for fGFPCA
# code prediction value at observed points as NA
df_est_latent$pred_t100[df_est_latent$bin <= 100] <- NA
df_est_latent$pred_t200[df_est_latent$bin <= 200] <- NA
df_est_latent$pred_t300[df_est_latent$bin <= 300] <- NA

```{r AUC_fGFPCA_Appl, cache=TRUE}
# maximum observations time
t_vec2 <- c(100, 200, 300)

# calculate AUC on unbinned grid
auc_mat2 <- matrix(NA, nrow = 3, ncol = 3) # row index iteration, col index max observation time

# result container for interpolation
Y_df2 <- df_appl %>% select(participant_id, starts_with("day")) %>% 
  pivot_longer(2:401, values_to = "Y", names_to = "sind_inx") %>%
  mutate(sind_inx = gsub("day_", "", sind_inx)) %>%
  mutate(sind_inx = as.numeric(sind_inx)) %>%
  rename("id" = "participant_id")
Y_df2$pred_100 <- Y_df2$pred_200 <- Y_df2$pred_300 <- NA

  # use interpolation to extend prediction to the unbinned grid
for(i in 1:N2){
   Y_df2$pred_100[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
                                         y=df_est_latent$pred_t100[df_est_latent$id==i], 
                                         xout = 1:J2)$y
   Y_df2$pred_200[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
                                         y=df_est_latent$pred_t200[df_est_latent$id==i], 
                                         xout = 1:J2)$y
   Y_df2$pred_300[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
                                         y=df_est_latent$pred_t300[df_est_latent$id==i], 
                                         xout = 1:J2)$y
  }

# split unobserved time into intervals of length 100
Y_df2$bin_cat <- cut(Y_df2$sind_inx, breaks=c(0, 100, 200, 300, 400), include.lowest = T)
Y_df_lst2 <- split(Y_df2, f = Y_df2$bin_cat)
Y_df_lst2 <- Y_df_lst2[-1]

for(j in seq_along(t_vec2)){

    t <- t_vec2[j]
    pred_name <- paste("pred_", t, sep = "")

    # calculate AUC
    auc_lst <- lapply(Y_df_lst2[j:3], 
                       function(x){
                         row_id <- !is.na(x[, pred_name])
                         this_eta <- x[row_id, pred_name]
                         this_Y <- x[row_id, "Y"]
                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
                         return(auc_pref@y.values)
                })

    auc_vec <- c(rep(NA, 3-length(auc_lst)), unlist(auc_lst))
    auc_mat2[ , j] <- auc_vec

colnames(auc_mat2) <- c("100", "200", "300")
rownames(auc_mat2) <- c("(100, 200]", "(200, 300]", "(300, 400]")

```{r AUC_GLMMadaptive_Appl}
# calculate AUC on unbinned grid
auc_mat_ref2 <- matrix(NA, nrow = 3, ncol = 3)  # row index iteration, col index max observation time

# split unobserved time into intervals of length 100
df_est_latent_ref$bin_cat <- cut(df_est_latent_ref$sind_inx, 
                                 breaks=c(0, 100, 200, 300, 400), include.lowest = T)
Y_df_lst2_ref <- split(df_est_latent_ref, f = df_est_latent_ref$bin_cat)
Y_df_lst2_ref <- Y_df_lst2_ref[-1]

for(j in seq_along(t_vec2)){

    t <- t_vec2[j]
    pred_name <- paste("pred_t", t, sep = "")

    # calculate AUC
    auc_lst <- lapply(Y_df_lst2_ref[j:3], 
                       function(x){
                         row_id <- !is.na(x[, pred_name])
                         this_eta <- x[row_id, pred_name]
                         this_Y <- x[row_id, "Y"]
                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
                         return(auc_pref@y.values)
                })

    auc_vec <- c(rep(NA, 3-length(auc_lst)), unlist(auc_lst))
    auc_mat_ref2[ , j] <- auc_vec

colnames(auc_mat_ref2) <- c("100", "200", "300")
rownames(auc_mat_ref2) <- c("(100, 200]", "(200, 300]", "(300, 400]")

```{r, Figure_appl, fig.cap="Predicted track of four randomly selected subjects from the daily weigh-in dataset. The grey solid line indicates the true latent continuous function. The dashed lines indicated predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
rand_id2 <- sample(N2, 4)
fig2a <- Y_df2 %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_100, pred_200, pred_300), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
  geom_line(aes(x=sind_inx, y = pred_100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Predicted latent function on probability scale", x = "Bin", 
       colour= "Observed time", title = "fGFPCA")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))

fig2b <- df_est_latent_ref %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t100, pred_t200, pred_t300), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
  geom_line(aes(x=sind_inx, y = pred_t100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Predicted latent function on probability scale", x = "Bin", 
       colour= "Observed time", title = "GLMMadaptive")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))

ggarrange(fig2a, fig2b, nrow = 1, common.legend = T)
```

```{r}
cbind(auc_mat2, auc_mat_ref2) %>%
  kable(booktabs = T, digit=2,
        caption = "AUC of fGFPCA and GLMM adaptive on the daily weigh-in dataset",
        align = "lcccccccc") %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "fGFPCA" = 3, "GLMMadaptive" = 3)) %>%
  add_header_above(c(" " = 1, "Maximum observed time" = 6)) %>% 
  group_rows(index = c("Prediction time window" = 3))

```
Discussion
=================================================================================================================================
-	Grid 
-	Score bias: cannot demonstrate without repeat simulation
